Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-02-25 papers/histogram/sad.tex
paper additions and figure formatting.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N50-lndos-comparison}
  \caption{The entropy of a square well fluid with 50 atoms and filling fraction
        0.3 as the green line.  The green hatched area reflects the
        minimum entropy change needed to converge to the true value.
        The light blue area is the quadratic approximation
        for the change in entropy.  The vertical dotted lines represent
        the energy corresponding to $T=1/3$ and $T=\infty$.}
        ~\jpsays{I fixed this to use samc-1e4 reference. Remove comment}
  \label{fig:entropy-cartoon}
\end{figure}

@@

@@

We use the average entropy error vs moves as a metric to compare
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a precise reference system. For each simulation, the reference system
is chosen to be the final output of a SAMC simulation with a fixed interesting energy range.
We note this because SAMC does not require an energy range as an input parameter; however,
by fixing the energy range to for SAMC, a smaller $t_0$ can be chosen considerably reducing the
simulation time.
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{ref}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a
different method take longer to converge. This behavior is seen in
Fig.~\ref{fig:N50-ff0.3-avg-error}.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  a fixed energy range SAMC $t_0 = 10^3$ serving as the reference.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-slow.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  a fixed energy range SAMC $t_0 = 10^3$ from Fig.~\ref{fig:N50-ff0.3-avg-error}
  serving as the reference.}\label{fig:N50-ff0.3-avg-error-slow}
\end{figure}

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average error in the
entropy as a function of time for this system with the reasonable
displacement distance of $\delta_0 = 0.05\sigma$.  The solid lines
represent the average of the absolute value of the error in the
entropy averaged over eight simulations using different random number
seeds.  The range of average errors for each simulation is shown as a
shaded region around its mean error.  By the time $10^8$ moves have
been made all but the SAMC simulation with the shortest $t_0$ have
begun to converge as $1/\sqrt{t}$.  We then see the WL error begin to
saturate around $10^{10}$ moves.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n50}
  \caption{\davidsays{We need to make a distinction between SAD and
      SAD both in figure and in caption.  I don't know what to call
      the slow SAD.  Also, should this move down to results, and do we
      want slow WL and slow-1/t WL?} The update factor $\gamma_t$ for
    $N=50$ versus iteration for three different methods: WL, SAMC, and
    SAD.}\label{fig:fast-slow-gamma}
\end{figure}

Figure~\ref{fig:N50-ff0.3-avg-error-slow} shows the average error in
the entropy as a function of time for this system with the
unreasonably small displacement distance of $\delta_0 = 0.005\sigma$.
%
The smaller translation scale causes all methods to take additional
time to explore all energies. Based on random walk scaling, a ideal
method should scale roughly as $\delta_0^{-2}$ in the limit of small
$\delta_0$, that is, one order of magnitude in the displacement
distance should result in two order of magnitude increase in
convergence time. SAMC simulations with a $t_0$ value that rapidly
converged for $\delta = 0.05\sigma$ do not converge at all for a
translation scale of $\delta = 0.005\sigma$. SAD, WL, and $1/t$-WL
handle the shift in displacement distance and converge roughly as
expected.

To see how SAD, WL, and $1/t$-WL compensate for the smaller
displacement distance, \davidsays{see Fig.~\ref{fig:fast-slow-gamma}
  and discuss this}

@@

\subsection{A periodic system with 256 atoms}
For this simulation, we chose a minimum reduced
temperature of $1.0$, which corresponds to an interesting energy range
from $-915$ to $-509$.  The number of important energy states for this
system is then $N_S = S_{\min} - S_{\max} = 406$.
The entropy of this system is has an entropy that
differs by 275, corresponding to a ratio of
$10^{119}$ between the highest and lowest density of states.

Figure~\ref{fig:N256} shows the average error in the
entropy as a function of time for this system with the reasonable
displacement distance of $\delta_0 = 0.05\sigma$.  The solid lines
represent the average of the absolute value of the error in the
entropy averaged over eight simulations using different random number
seeds.  The range of average errors for each simulation is shown as a
shaded region around its mean error.  By the time $10^8$ moves have
been made all but the SAMC simulation with the shortest $t_0$ have
begun to converge as $1/\sqrt{t}$.  We then see the WL error begin to
saturate around $10^{10}$ moves.

@@

{\color{red}
Figure~\ref{fig:n256-lndos-comparison} shows the calculated entropy for
the $N = 256$ system. The grey dash lines indicate the $S_{\max} = 509$
point and the $S_{\min} = 915$. Only the reference method is shown.
}

>>>>

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N50-lndos-comparison}
  \caption{The entropy of a square well fluid with 50 atoms and filling fraction
        0.3 as the green line.  The green hatched area reflects the
        minimum entropy change needed to converge to the true value.
        The light blue area is the quadratic approximation
        for the change in entropy.  The vertical dotted lines represent
        the energy corresponding to $T=1/3$ and $T=\infty$.}
        ~\jpsays{I fixed this to use samc-1e4-s1. Remove comment}
  \label{fig:entropy-cartoon}
\end{figure}

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The average entropy error for each MC method is shown for $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$ with
               a fixed energy range reference as discussed in the text. The update factor $\gamma_t$ for
               $N=50$ versus iteration for the four different methods: WL, $1/t$-WL, SAMC, and SAD is shown beneath.
           (b) The average entropy error for each MC method is shown for $\delta_0 = 0.005\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$ with
               the same fixed energy range reference used in (a). The update factor $\gamma_t$ for
               $N=50$ versus iteration for the four different methods: WL, $1/t$-WL, SAMC, and SAD is shown beneath.}\label{fig:fast-slow-gamma}
\end{figure*}

@@

We use the average entropy error vs moves as a metric to compare
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a precise reference system. For each simulation, the reference system
is chosen to be the final output of a SAMC simulation with a fixed interesting energy range.
We note this because SAMC does not require an energy range as an input parameter; however,
by fixing the energy range to for SAMC, a smaller $t_0$ can be chosen considerably reducing the
simulation time.
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{ref}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a
different method take longer to converge. This behavior is seen in
Fig.~\ref{fig:fast-slow-gamma}.

@@

%\begin{figure}
%  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-default.pdf}
%  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
%  a fixed energy range SAMC $t_0 = 10^3$ serving as the reference.}\label{fig:N50-ff0.3-avg-error}
%\end{figure}
%\begin{figure}
%  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-slow.pdf}
%  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
%  a fixed energy range SAMC $t_0 = 10^3$ from Fig.~\ref{fig:N50-ff0.3-avg-error}
%  serving as the reference.}\label{fig:N50-ff0.3-avg-error-slow}
%\end{figure}

Figure~\ref{fig:fast-slow-gamma}a shows the average error in the
entropy as a function of time for this system with the reasonable
displacement distance of $\delta_0 = 0.05\sigma$.  The solid lines
represent the average of the absolute value of the error in the
entropy averaged over eight simulations using different random number
seeds.  The range of average errors for each simulation is shown as a
shaded region around its mean error.  By the time $10^8$ moves have
been made all but the SAMC simulation with the shortest $t_0$ have
begun to converge as $1/\sqrt{t}$.  We then see the WL error begin to
saturate around $10^{10}$ moves.

Figure~\ref{fig:fast-slow-gamma}b shows the average error in
the entropy as a function of time for this system with the
unreasonably small displacement distance of $\delta_0 = 0.005\sigma$.
%
The smaller translation scale causes all methods to take additional
time to explore all energies. Based on random walk scaling, a ideal
method should scale roughly as $\delta_0^{-2}$ in the limit of small
$\delta_0$, that is, one order of magnitude in the displacement
distance should result in two order of magnitude increase in
convergence time. SAMC simulations with a $t_0$ value that rapidly
converged for $\delta = 0.05\sigma$ do not converge at all for a
translation scale of $\delta = 0.005\sigma$. SAD, WL, and $1/t$-WL
handle the shift in displacement distance and converge roughly as
expected.

The methods SAD, WL, and $1/t$-WL compensate for the smaller
displacement distance as can be seen from Figure~\ref{fig:fast-slow-gamma}.
The update factors take approximately 10x longer to reach steady-state for
the slower translation scale. Because of this update behavior, these methods
do not suffer in converging to the correct density of states from having a
static parameter such as SAMC $t_0$ in the update factor. SAMC methods that converge
in a reasonable amount of moves in Figure~\ref{fig:fast-slow-gamma}a do not 
necessarily do so for Figure~\ref{fig:fast-slow-gamma}b.

Sad performs similarly regardless of the displacement distance chosen although
it transitions to $1/\sqrt{t}$ convergence after

@@

\subsection{A periodic system with 256 atoms}
For this simulation, we chose a minimum reduced
temperature of $1.0$, which corresponds to an interesting energy range
from $-915$ to $-509$.  The number of important energy states for this
system is then $N_S = E_{\min} - E_{\max} = 406$.
The entropy of this system is has an entropy that
differs by 275, corresponding to a ratio of
$10^{119}$ between the highest and lowest density of states.

Figure~\ref{fig:n256} shows the average error in the
entropy as a function of moves for this system with the reasonable
displacement distance of $\delta_0 = 0.05\sigma$.  The solid lines
represent the average of the absolute value of the error in the
entropy averaged over eight simulations using different random number
seeds.  The range of average errors for each simulation is shown as a
shaded region around its mean error.  By the time $10^8$ moves have
been made all but the SAMC simulation with the shortest $t_0$ have
begun to converge as $1/\sqrt{t}$.  We then see the WL error begin to
saturate around $10^{10}$ moves.

@@

%{\color{red}
%Figure~\ref{fig:n256-lndos-comparison} shows the calculated entropy for
%the $N = 256$ system. The grey dash lines indicate the $S_{\max} = 509$
%point and the $S_{\min} = 915$. Only the reference method is shown.
%}
